---
title: "Probability in Biomedical Data Science"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"
format: pptx

execute:
  echo: true
  warning: false
  message: false
---
```{r, echo=FALSE}
library(tidyverse)
```


## From Sampling to Probability

<div style="font-size:0.7em">

In the previous lecture, we introduced four statistical concepts:

- **population** — the complete set of units or outcomes of interest  
- **sample** — the observed subset drawn from the population  
- **parameter** — a numerical characteristic of the population  
- **statistic** — a numerical summary computed from the sample  

Probability provides the framework that explains:

- why samples vary  
- how statistics fluctuate across repeated samples  
- how parameters can be inferred from observed data  

</div>

---

## Biomedical Data Analysis and Probability

<div style="font-size:0.7em">

Biomedical systems exhibit inherent variability:

- patients differ biologically and environmentally  
- measurements contain technical variation  
- disease processes evolve stochastically  

- probability is a framework that can be used to describe this variability
- biomedical data analysis often begins by specifying probabilistic models for outcomes

Probability models allow researchers to:

- represent uncertainty
- describe how data arise  
- quantify expected variability across individuals  


</div>
---

## Probability: Defnition

<div style="font-size:0.7em">


- a measure of how likely an event is to occur  
- interpreted through long-run relative frequency under repeated trials  

- the mathematical representation of uncertainty in data-generating processes  

Formally:

$$
0 \le P(A) \le 1
$$

Probability assigns a numerical measure to uncertainty about an event, where:

- 0 : the event does not occur  
- 1 : the event will occur with some degree of certainty   
- intermediate values quantify uncertainty  
- values between 0 and 1 quantify the degree of uncertainty 

Biomedical examples:

- Probability a patient responds to treatment
- Probability a variant is observed in a population
- Probability of disease occurrence


</div>

---

## Probability and Parameters

<div style="font-size:0.7em">

Population parameters often represent probabilities

Examples:

- mutation prevalence in a population  
- treatment response probability  
- disease incidence risk  

Parameters describe the population probability distribution

Because parameters are unobserved:

- samples are collected  
- statistics are computed  
- probability explains variability  


</div>

---

## Probability and Statistics

<div style="font-size:0.7em">

The relationship between probability and statistics can be summarized as:

- **probability** — describes population behavior and data generation  
- **statistics** — uses samples to learn about probability structure  

Conceptually:

$$
\text{Probability} \rightarrow \text{Sampling variability} \rightarrow \text{Statistical inference}
$$

Probabiliy and statistics are linked to the following:

- sampling variability arises from probabilistic selection 
- models assume outcomes follow probability distributions  
- workflow begins with specifying outcome distribution  

</div>

---

## Probability in Biomedicine

Biological systems are variable

Even under identical conditions:

- Patients respond differently to treatment  
- Cells express genes differently  
- Disease progression varies  

Probability provides a framework to model this variability

---

### Transition to Formal Probability Concepts

<div style="font-size:0.7em">

Having established probability as the mechanism linking population and sample, we now introduce formal components of probability models:

- events and probability assignments  
- random variables  
- probability distributions  
- discrete and continuous outcome structures  

These objects allow biological uncertainty to be expressed mathematically and estimated from data.

</div>

---

### Random Variables: Connecting Outcomes to Numbers
<div style="font-size:0.7em">
**random variable**: 

- assigns a numerical value to the outcome of a random process
- repreent biological outcomes or values in the form of analyzable quantities
- numerical representation of an outcome whose value is determined by a probabilistic process



</div>
---

## Random Variables: Representing Biomedical Outcomes

<div style="font-size:0.7em">

- random variables assign numerical values to outcomes generated by probabilistic processes

- random variables define the outcome layer of statistical models and serve as the objects whose distributions are studied.

Examples:

- mutation presence indicator  
- treatment response category  
- biomarker measurement value  
- survival time  


</div>

---

### Random variable types in biomedicine

Binary outcome:

$$
Y =
\begin{cases}
1 & \text{mutation present}\\
0 & \text{mutation absent}
\end{cases}
$$

Categorical outcome:

$$
Y \in \{\text{CR}, \text{PR}, \text{NR}\}
$$

Continuous outcome:

$$
Y = \text{tumor size}
$$

---

### Binary random variables

Binary outcomes often represent presence/absence decisions

Examples:

- pathogenic variant present vs absent
- treatment response (yes/no)
- adverse event (yes/no)

---

### Code example: simulate a binary mutation indicator
<div style="font-size:0.7em">

**Biomedical question**

What proportion of patients in a population carry a pathogenic mutation in a gene of interest?

**Population perspective**

- each patient either carries the mutation (1) or does not (0)  
- the population mutation prevalence is an unknown probability parameter \(p\)  
- this setting is naturally modeled with a **Bernoulli distribution**

Thus, for patient \(i\):

$$
Y_i \sim \text{Bernoulli}(p)
$$

**Sampling and estimation**

- a study observes \(n\) patients drawn from this population  
- each observation is a Bernoulli trial  
- the sample proportion estimates the population mutation probability  

</div>
---

### Code example: Binary Mutation Indicator
<div style="font-size:0.6em">

Explanation:

- `rbinom(n, 1, p)` generates \(n\) independent Bernoulli outcomes  
- mimics sampling patients from the population distribution 
- each simulated patient outcome represents a draw from the population distribution, and the sample proportion summarizes these draws to estimate the population probability

Example:

- a study observes *n* patients drawn from this population  
- each patient outcome is a Bernoulli trial (mutation present or absent)  
- the table summarizes observed counts and proportions across sampled patients  

```{r, echo=FALSE}
#| label: l2-binary-sim
n <- 200
p <- 0.17
mutation <- rbinom(n, size = 1, prob = p)

table <- tibble(mutation) |> count(mutation) |> mutate(prop = n / sum(n))
```


```{r}
table
```

Interpretation:

mutation = 0 → 162 patients do not carry the mutation [81% of the sample]

mutation = 1 → 38 patients carry the mutation [19% of the sample]

</div>
---

### Categorical random variables

<div style="font-size:0.6em">

Categorical outcomes represent membership in one of several classes

Examples:

- tumor subtype [LumA, LumB, HER2, Basal]
- response category [CR/PR/NR]
- lineage assignment in single-cell data

</div>
---

### Code example:Clinical Trial Patient Responses

<div style="font-size:0.6em">

**Biomedical question**

In a clinical trial evaluating a new therapy, patients are classified into one of three mutually exclusive response categories:

- **CR** — complete response  
- **PR** — partial response  
- **NR** — no response  

**Population perspective**

Across all eligible patients, treatment response follows a categorical probability distribution with unknown probabilities:

$$
P(Y=\text{CR}) = p_{CR}, \quad
P(Y=\text{PR}) = p_{PR}, \quad
P(Y=\text{NR}) = p_{NR}
$$

- probabilities describe the **population response distribution**
</div>
---


### Code example:Clinical Trial Patient Responses

<div style="font-size:0.5em">


**Sampling and estimation**

- a study observes \(n\) patients drawn from this population  
- each patient outcome represents one draw from the categorical distribution  
- sample counts and proportions estimate the underlying response probabilities  

```{r, echo=TRUE}
#| label: l2-cat-sim
n <- 300
levels_resp <- c("CR", "PR", "NR")
probs_resp  <- c(0.20, 0.50, 0.30)

response <- sample(levels_resp, size = n, replace = TRUE, prob = probs_resp)

tab <- tibble(response = factor(response, levels = levels_resp)) |>
  count(response) |>
  mutate(prop = n / sum(n))

tab
```

Interpretatioin:

For each response category:

- **CR, PR, NR** : possible treatment outcomes  
- **n** : number of patients observed in that category [sample counts] 
- **prop** : proportion of patients in that category [sample probabilities]  

Treatment response outcomes among the 300 sampled patients:

Complete Response[CR]: 68 patients representing 22.7% of the sample

Partial Response [PR]: 155 patients representing 51.7% of the sample

No Response [NR]: 77 patients representing 25.7% of the sample

</div>
---


## Continuous random variables

- continuous outcomes arise from measurements on a scale
- modeled with densities
- associated models: linear regression and survival modeling

Examples:

- biomarker concentration
- tumor size
- expression level (after suitable transformation)



---

### Code example: Continuous Biomarker
<div style="font-size:0.6em">

**Biomedical context**

- circulating tumor biomarker concentrations vary continuously across patients due to biological heterogeneity and measurement variability  
- measurements arise from a population distribution characterized by a mean level and variability

- `rnorm()`: simulates draws from a normal population distribution
```r
biomarker <- rnorm(n, mean = mu, sd = sigma)
```

- each observation represents a sampled patient measurement
- histogram visualizes the empirical distribution observed in the sample


```{r, echo=FALSE}
#| label: l2-cont-sim2
n <- 400
mu <- 10
sigma <- 2

biomarker <- rnorm(n, mean = mu, sd = sigma)

plot_cont <- ggplot(tibble(biomarker), aes(biomarker)) +
  geom_histogram(bins = 35) +
  labs(x = "Biomarker", y = "Count",
       title = "Simulated biomarker measurements")
```




```{r}
#| fig-width: 7
#| fig-height: 4
plot_cont
```

</div>
---

### Code example: Continuous Biomarker
<div style="font-size:0.6em">


Explanation:

- `rnorm()` simulates a continuous variable under a normal model
- biomarker concentration is the continuous random variable
- histogram shows the distribution of observed values in the sample
- each simulated value represents one patient biomarker measurement drawn from the population distribution
- histogram shows how these measurements are distributed across the cohort


```{r, echo=FALSE}
#| label: l2-cont-sim
n <- 400
mu <- 10
sigma <- 2

biomarker <- rnorm(n, mean = mu, sd = sigma)

plot_cont <- ggplot(tibble(biomarker), aes(biomarker)) +
  geom_histogram(bins = 35) +
  labs(x = "Biomarker", y = "Count",
       title = "Simulated biomarker measurements")
```

```{r}
#| fig-width: 7
#| fig-height: 4
plot_cont
```


</div>
---

### Population Distributions as Models of Biological Variation

<div style="font-size:0.7em">

- `population distribution`: describes how outcomes vary across all individuals in a population
- distributions represent the probabilistic structure of the population
- probability distributions specify:

  - the possible values of a random variable  
  - the probability associated with each value  

In a biomedical context, distributions characterize:

  - variability across patients  
  - heterogeneity across cells  
  - variation across populations  

- statistical inference aims to learn these distributions from sampled observations

</div>

---

### Probability Distributions
<div style="font-size:0.7em">
A probability distribution specifies:

  - represents the population and describe population behavior
  - probability of each possible value of a random variable
  - describes the behavior of the population

Examples:

- distribution of treatment response
- distribution of biomarker levels
- distribution of gene expression

</div>
---


### Why Distributions Are Fundamental to Statistics
<div style="font-size:0.6em">

- statistical models assume observed data arise from an underlying probability distribution
- probability distributions: provide the mathematical representation of biological variability across individuals, cells, and populations

- distribution defines:

  - possible outcomes  
  - how frequently those outcomes occur  
  - variability expected across individuals  

**Examples of outcome distributions in biomedical models**

- **logistic regression**: assumes a **Bernoulli distribution** for binary outcomes
  [i.e.: response vs no response, mutation present vs absent]

- **multinomial regression**: assumes a **categorical/multinomial distribution**  
  [i.e.: complete, partial, non-response categories, tumor subtype]

- **survival analysis**: assumes a **time-to-event distribution**  
  [i.e.: exponential, Weibull, or Cox hazard-based models for survival time]

- **variant frequency estimation**: commonly assumes **binomial or multinomial distributions**  
  [i.e.: allele counts, genotype counts in populations]


</div>
---

## Probability Distributions Describe Population Behavior

<div style="font-size:0.7em">

- **population distribution**: describes how outcomes vary across all individuals or units in a population

- probability distribution specifies:

  - the set of possible values a random variable can take  
  - the probability associated with each value (or range of values)  

- distibutions represent the probability structure of the population

Example Interpretation:

- study first collects individual observations
- samples provide partial views of the population
- distribution describes long-run population behavior  

Distributions vs Individual Observations

  - individual observation: one patient tumor size
  - distribution: all possible tumor sizes and their probabilities


</div>

---

## Probability Distributions Describe Population Behavior

<div style="font-size:0.7em">


**Biomedical examples**

- probability of treatment response categories [i.e.: complete, partial, nonresponse]  
- distribution of biomarker concentrations across patients  
- distribution of gene expression levels across cells  
- prevalence distribution of pathogenic variants in a population  

**Relevance in biomedical research**

Population distributions can be used to:

- quantify biological variability across individuals and systems  
- define parameters of interest [means, proportions, rates, risks]
- determine expected sampling variability  

- statistical approaches aim to learn the population distribution from sample data in order to make inferences about samples

</div>

---

### Summary

<div style="font-size:0.7em">

Key concepts introduced:

- probability quantifies biological uncertainty and data generation  
- random variables represent biomedical outcomes numerically  
- probability distributions describe population behavior and variability  
- sampling produces realizations from population distributions  
- statistics summarize sampled outcomes  
- inference uses samples to learn about underlying distributions  

- concepts collectively establish the probabilistic foundation for regression, survival analysis, and biomedical data science

</div>

---

