---
title: "Sampling Distributions, Sample Size, and Statistical Power"
author: "<br><br><span style='font-size:25px;'><strong>Lesley Chapman Hannah, Ph.D., M.S.</strong></span><br>College of Graduate Studies<br>Northeast Ohio Medical University"
format:
  revealjs:
    theme: simple
    slide-number: true
#    incremental: true
    chalkboard: true
    transition: fade
execute:
  echo: true
  warning: false
  message: false
---

```{r, echo=FALSE}
#| label: setup-lecture3
library(tidyverse)
set.seed(29)
```


```{r, echo=FALSE}

library(tidyverse)
set.seed(29)

# synthetic population
N <- 200000
pop <- tibble(
  age_group = sample(c("Younger", "Older"), size = N, replace = TRUE, prob = c(0.6, 0.4)),
  # biomarker depends on age group (e.g., systolic BP, CRP, etc.)
  biomarker = if_else(age_group == "Older",
                      rnorm(N, mean = 130, sd = 15),
                      rnorm(N, mean = 118, sd = 15))
)

```

### Lecture Overview
<div style="font-size:0.6em">
**Sampling distributions**

- how statistics vary across repeated samples

**Sampling strategies**

- procedures used to select observations from a population

**Sample size and uncertainty**

- how the number of observations affects precision and power

**Precision and statistical power**

- designing studies that can estimate parameters or detect meaningful effects
</div>
---

## Sampling distribution: Definition

- **sampling distribution**: probability distribution of a statistic computed from all possible samples of a fixed size from a population

Key distinction(s):

- population parameter: fixed, unknown constant [μ, p, λ]
- sample statistic: random variable [$\bar{X}$, $\hat{p}$, etc.]
- sampling distribution: distribution of the statistic across repeated samples


---

### Sampling in biomedical research
<div style="font-size:0.6em">

**Review**: we rarely observe an entire population [i.e.: all patients, all tumors, all genomes] instead a sample is taken and generalized for statistical analysis

- sampling, bias, and sample size determine whether your study’s conclusions are:

  - valid [not systematically wrong]
  - precise [narrow uncertainty]
  - reproducible [another cohort gives compatible results]
  - actionable [effects are detectable at clinically meaningful scales]

</div>

---

### Sampling and Statistical Analysis
<div style="font-size:0.6em">

A core objective of statistical analysis is to determine:

> Does this sample allow the researcher to learn something true about a population?


Three components determine the answer:

1. **sampling strategy**: defines who your data stand for
2. **bias**: determines whether your estimate is systematically distorted.
3. **sample size**: determines how much random uncertainty remains


</div>

---

### Sampling strategies

<div style="font-size:0.6em">

**Sampling strategy**: 

  - pre-specified procedure for selecting observations from a defined target population 
  - determines who can be selected, how selection occurs, and each unit’s inclusion probability

**Importance in practice**:

- determines whether estimates [i.e.: prevalence, mean biomarker level, incidence rate] generalize to the target population
- determines whether standard errors / confidence intervals are trustworthy
- determines the feasibility and cost of data collection [i.e.: multi-site, registries, EHRs, biobanks]

**Examples**:

- selecting patients from a hospital registry to estimate readmission risk
- selecting biospecimens from a biobank to characterize a biomarker distribution
- selecting sequencing samples from a case series to study mutation spectra

</div>

---

### Sampling Strategies
<div style="font-size:0.6em">

- **simple Random Sample [SRS]**: every set of n units has equal chance of selection
  - strength: clean inference
  - weakness: requires full sampling frame

- **stratified sampling**: divide into strata [age, sex, ancestry, site] then sample within strata

  - strength: ensures subgroup representation and increases precision
  - note: disproportionate sampling may require weights

- **cluster/multistage sampling**: sample clusters [hospitals/clinics] then sample individuals within clusters

- **systematic sampling**: random start, then every $k^{th}$ unit from an ordered list

</div>

---


### Sample size considerations
<div style="font-size:0.6em">

**sample size considerations**: 

  - structured decisions used to determine how many observational units are required within a sample
  - how the observations should be distributed across groups or strata so that statistical estimates are sufficiently precise and comparisons have adequate power
  
structured process used to determine if there is enough data to answer the scientific question reliably, and considers:

  - how many observational units [patients, specimens, encounters, variants] must be included
  - how those units should be allocated across groups or strata


sample size is determined by several factors such as:

  - parameter you want to estimate [mean, proportion, rate, hazard]
  - variability of the outcome
  - smallest clinically meaningful effect
  - sampling design [SRS, stratified, clustered]




</div>

---

### Sample size considerations
<div style="font-size:0.6em">

**Small sample size characteristics**:

  - estimates are more variable
  - standard errors are larger
  - confidence intervals are wider
  - statistical tests have lower power
  - sampling distributions may not approximate normality

**Implications**:

  - results could be less stable across repeated samples
  - effect estimates may fluctuate substantially
  - type II error [missing a real effect] is more likely

**Commonly found in**:

  - rare disease studies
  - early-phase clinical trials
  - pilot studies

</div>

---


### Sample size considerations
<div style="font-size:0.6em">

**large sample sizes**:

  - estimates are more precise
  - smaller standard errors
  - narrower confidence intervals
  - higher statistical power
  - sampling distributions approach normality 

**implications**:

  - greater stability across repeated sampling
  - smaller effects become detectable
  - more reliable inference

</div>

---

### Why Sample Size Matters

Data is collected to answer two types of questions:

1. **Estimation questions**  
   - What is the value of a biological parameter?

2. **Comparison questions**  
   - Do groups differ in a meaningful way?

- sample size planning helps ensure that the data collected contain **enough information** to answer the research question

---


### Sample Size in Practice
<div style="font-size:0.6em">

1) `precision`: focuses on **estimation accuracy**

  - Goal: choose sample size so that a confidence interval is narrow enough for the research question
  - Core question: How uncertain is our estimate?
  - larger n → smaller standard error → narrower confidence interval → greater precision
  - example: estimating a treatment response rate within $±3%$
  
Example:

> Estimate the response rate within **±3%**

- sample size is chosen so the confidence interval width approximates the true estimate

Interpretation:

- estimate will vary across samples, but with sufficient $n$ the uncertainty becomes small enough to answer the research question


</div>

---

### Sample Size in Practice
<div style="font-size:0.6em">

2)**Power**: detecting differences between groups

  - goal is to compare groups or test a hypothesis
  - Core question: If the true effect exists, what is the probability we detect it?
  - choose n so that you can detect a clinically meaningful effect with high probability
  - answers: If the true effect exists, what is the probability we detect it?
  
Depends on factors such as:
  
  - effect size: smaller effects require larger n
  
  - variability: more variability requires larger n
  
  - outcome type: binary, continuous, count, survival
  
</div>

---
  
### Sample Size in Practice
  <div style="font-size:0.6em">
  
  
  3) **design adjustments**: : adjustments that ensure the study still contains enough effective information
  
  - design adjustments correct for:
  - **independent observations**
  - participants are naturally grouped [i.e.: patients grouped by hospital] > observations within a group/cluster tend to be correlated >> correlated observations contribute less independent information and reduces the effective sample size
- solution: increase total enrollment to compensate
- **complete data**: data missingness
- loss of data reduces precision [wider confidence intervals] and power [lower probability of detecting true effects]
- **simple random sampling**
  - helps define the statistical benchmark for how information accumulates with increasing n

</div>
  
---


### Sample Size and Power Calculations
  
<div style="font-size:0.6em">

Note:
    
> **Sample size depends on the inferential goal.**
    
Before computing our sample size (n), consider:
    
1) **Precision [estimation goal]**: degree to which repeated samples produce similar estimates of a population parameter

Estimate a population parameter with sufficient precision

Precision increases when:
  
- sample size [n] increases
- outcome variability decreases
- measurement error decreases

</div>
  
---
  
### Sample Size and Power Calculations
  
<div style="font-size:0.5em">
Note:
  
> **Sample size depends on the inferential goal.**
  
Before computing our sample size (n), consider:
  
2) **Power [comparison goal]** : probability that a statistical test correctly rejects a false null hypothesis  

Detect a meaningful biological difference with high probability

Power depends on:
  
- effect size [magnitude of true difference]
- variability
- sample size
- significance level

Note:
  
- precision : how well we measure
- power : reliably we detect a signal
</div>
  
---


### Sample Size Calculations

Before collecting data, we must answer:

1. How precise do we want our estimates to be?
2. How likely do we want to detect a meaningful difference?

These are two different goals:

- **precision** [estimation]
- **power** [hypothesis testing]

---

### Sample Size and Power Calculation: Example
  
<div style="font-size:0.6em">
  
**Background**:  

  - Tumor Mutational Burden (TMB) is the number of somatic mutations per megabase (mut/Mb)
  - In colorectal cancer, tumors with deficient mismatch repair (**dMMR**) often show higher TMB than proficient mismatch repair (**pMMR**) tumors
  
**Research Task**:
- A colorectal cancer genomics team is deciding whether to invest in deeper sequencing for mismatch-repair testing
- The group wants to know whether dMMR tumors have meaningfully higher Tumor Mutational Burden (TMB) than pMMR tumors, using TMB as a quantitative signal of DNA repair deficiency


**Why this is a good power calculation example**:
  
- outcome is **continuous** (mut/Mb), typical of genomics summary measures  
- the effect size can be stated in biologically meaningful units (mut/Mb)

**Power/Sample Size Questions**:

`Precision question [determine sample size needed to estimate parameter]`:

- How many dMMR tumor samples do we need so that our estimate of the mean TMB has a 95% confidence interval no wider than a prespecified margin [i.e.: ±2 mut/Mb]?

`Power question [determine power to detect signal with sample size]`:

How many tumors per group do we need so that we have 80% power to detect it at α = 0.05?

</div>
  
---

### Key Definitions
<div style="font-size:0.5em">
**parameter**: fixed but unknown population value [i.e.: $\mu_{dMMR}$]

**statistic**: value calculated from the sample [i.e.: $\bar{x}_{dMMR}$]

**standard error [SE]**: variability of a statistic across repeated samples

**α**: false positive rate 

  - Type I error - rejecting the null hypothesis when it is actually true
  - significance level is set to a fixed value: α=0.05
  
**power = 1 − β**: 

  - probability that your study will correctly detect a real effect
  - β = probability of a Type II error [missing a real effect]

For a mean:
$$
SE = \frac{\sigma}{\sqrt{n}}
$$

- $\sigma$ = population standard deviation  
- $n$ = sample size  
- $n$ increases $\rightarrow$ SE decreases
</div>
---


### Example: Estimating Mean TMB

<div style="font-size:0.5em">


The following determines the sample size needed to estimate a mean given a desired **margin of error**

margin of error for a confidence interval is:

$$
ME = z \cdot \frac{\sigma}{\sqrt{n}}
$$

Determine the following:

Define the variability

- $\sigma = 10$ mut/Mb  
  - **σ**: estimated standard deviation of TMB from prior studies (mut/Mb)



</div>

---


### Example: Estimating Mean TMB

<div style="font-size:0.5em">


Define margin of error

- Desired margin of error = 2 mut/Mb  

Confidence Level [z-score]:

- 95% confidence: $z = 1.96$

Sample size formula:

- calculate the minimum number of tumors required

$$
\left(\frac{z \sigma}{ME}\right)^2
$$

</div>

---

### Example: Estimating Mean TMB

<div style="font-size:0.5em">


```{r}
sigma <- 10
ME <- 2
z <- 1.96

n_precision <- (z * sigma / ME)^2

#round up to ensure adequate precision
ceiling(n_precision)
```


Interpretation:

- study needs **at least 97 tumors** to estimate the mean TMB with a 95% confidence interval whose half-width is ±2 mut/Mb

- if the study needs to be repeated study many times the estimated mean TMB would typically fall within 2 mut/Mb of the true population mean

Note:

Increasing sample size reduces uncertainty because

$$
SE = \frac{\sigma}{\sqrt{n}}
$$

As $n$ increases:

- standard error decreases
- confidence intervals become narrower
- estimates become more precise

</div>

---

### Summary

- sampling distributions describe how a statistic behaves across repeated samples and provide the foundation for statistical inference. 

- sampling design determines validity

  - who the sample represents

  - whether estimates generalize to the population. 

- sample size determines uncertainty

  - larger samples → smaller standard errors

  - narrower confidence intervals

  - greater statistical power

---

